---
title: "Data Import"
output: github_document
---

This file is for doing data import. 

```{r message = F}
library(tidyverse)
library(readxl)
```

How to specify where a file is: 
* absolute path: start far away, line by line, say where exactly the file is
* relative path: relative to where you are right now, where is file 
  * always use relative paths 

# Import our data 

Import our first data set. 

```{r}
litters_df = 
  read_csv(file = "data_import_examples/FAS_litters.csv", 
           na = c("", ".", "NA")
           )

litters_df = janitor::clean_names(litters_df)
```
Go back to look at OG data set, they used both NA and . to indicate missing values. First, use janitor::clean_names(df) to clean up data names. Janitor package - don't need to load it all with library, you just use this function so you can just load the clean names function. 

Import your second dataset. 

```{r}
pups_df = 
  read_csv(file = "data_import_examples/FAS_pups.csv", 
           skip = 3, 
           na = c("", ".", "NA")
        )

pups_df = janitor::clean_names(pups_df) 
```

Can also skip writing file at beginning of parentheses (it expects this). Skip importing first three lines of data from pups dataset, because they're blank. 

Since you're setting the janitor function equal to the df, you're permanently altering the file. 

# Look at the data 

```{r}
litters_df
pups_df
```

Head or tail 

```{r}
head(litters_df, 20)

tail(pups_df, 5)
```
Looking at specified number of rows at head or tail of dataset. 

Skimming?

```{r}
skimr::skim(pups_df)
```
This skims the file and give you an overview. 
Saying `view(name of dataset)` in console will help you view dataset in R. Don't put in rmd because when you knit, you'll get a pop up window. 

Don't use read.csv, only use read_csv (the new and improved version) 

look at ?read_csv, you'll see what you can do within the function, including na = c("", "NA"), (that excludes the data that is blank or says NA)

# Fixing the missing data in dataframes 
Just go back to OG load and add line of `na = c("", ".", "NA")` and then re-run. This tells R to ignore those values (treat them as a missing value), rather than making the variable a character value. 

# Oh excel... Importing from Excel 
Go back to top, and load the readxl library. 
Jenny Bryan made `readxl` solve our problems. 

```{r}
mlb_df = 
  read_excel("data_import_examples/mlb11.xlsx")
```

Look at the data 

```{r}
mlb_df
```
Can use ?read_excel to specify stuff. Can specify which sheet, which cell range. 

Load the LotR data. 

Import FOTR words. 

```{r}
fotr_df =
  read_excel("data_import_examples/LotR_Words.xlsx",
             range = "B3:D6"
             )

fotr_df
```

Range says which cells to include. Don't include title. You can include sheet and range togeter (see help file) 

Same for Two Towers:
```{r}
tt_df =
  read_excel("data_import_examples/LotR_Words.xlsx",
             range = "F3:H6"
             )

tt_df
```

If you open a microsoft doc (Excel or word) while in repo, it will ask you to commit. Don't! Just close the file and it will go away. 


