---
title: "Data Manipulation"
author: "Emma"
date: "2026-09-24"
output: github_document
---

This file is for doing data manipulation. 

```{r message = F}
library(tidyverse)
```

Load the litters and pups datasets. 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", 
           na = c("", "NA", "."))

litters_df = janitor::clean_names(litters_df)

pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", 
           skip = 3,
           na = c("", "NA", "."))

pups_df = janitor::clean_names(pups_df)
```

## Select some variables 
Which data frame, and which variable columns are you selecting 
can list the columns, or do this through that (:), the - will do everything except what you -
GROUP = group renames that variable by itself (will also put it first), the everything says keep everything else

```{r}
select(litters_df, group, litter_number)

select(litters_df, group:gd_of_birth)

select(litters_df, -pups_survive)

select(litters_df, -group)

select(litters_df, starts_with("gd"))

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP = group, everything())

select(litters_df, -starts_with("gd"))
```

A couple alternatives 
```{r}
rename(litters_df, GROUP = group)
#Same thing as above except you don't have to tell it to keep everything else 

relocate(litters_df, litter_number, gd0_weight)
# relocate moves litter_number first, gd0_weight second (based off list order)
```


In the pups data, select litter number, sex, and PD ears: 

```{r}
select(pups_df, litter_number, sex, pd_ears)
```


