---
title: "Data Manipulation"
author: "Emma"
date: "2026-09-24"
output: github_document
---

This file is for doing data manipulation. 

```{r message = F}
library(tidyverse)
```

Load the litters and pups datasets. 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", 
           na = c("", "NA", "."))

litters_df = janitor::clean_names(litters_df)

pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", 
           skip = 3,
           na = c("", "NA", "."))

pups_df = janitor::clean_names(pups_df)
```

## Select some variables 
Which data frame, and which variable columns are you selecting 
GROUP = group renames that variable by itself (will also put it first), the everything says keep everything else

```{r}
select(litters_df, group, litter_number)
# can list the columns
select(litters_df, group:gd_of_birth)
# or do this through that (:)
select(litters_df, -pups_survive)
# - will do everything except what you -
select(litters_df, -group)

select(litters_df, starts_with("gd"))

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP = group, everything())

select(litters_df, -starts_with("gd"))
```

A couple alternatives 
```{r}
rename(litters_df, GROUP = group)
#Same thing as above except you don't have to tell it to keep everything else 

relocate(litters_df, litter_number, gd0_weight)
# relocate moves litter_number first, gd0_weight second (based off list order)
```


In the pups data, select litter number, sex, and PD ears: 

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## Filter some rows 
Which rows do I need to keep? Filter and keep all the litters where x, y (something inside your data set)
```{r}
filter(litters_df, gd_of_birth == 19)
# Asking a question with ==, which of gd_of_birth is equal to 20? 

filter(litters_df, pups_born_alive <= 5)
# less than or equal to 5 

filter(litters_df, pups_born_alive != 6)
# not equal to 6 (gets rid of 6)

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive < 7)
# going to keep both of those two options only, can filter for multiple things at same time

filter(litters_df, pups_born_alive < 4  | pups_born_alive > 8)
#this means or |

drop_na(litters_df)
# gets rid of all rows you have missing value 

drop_na(litters_df, gd0_weight)
# drop na for all in this specific column
```

In the pups dataset, keep only rows where PD walk is less than 11 and sex is 2 

```{r}
filter(pups_df, pd_walk < 11, sex == 2)
```

#Mutate to change some columns 

Let's add and change some columns 

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)
# creates a column in dataset, what is weight gain from beginning of pregnancy to right before birth
# puts variable at tail end
# does not change anything else 

mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight, 
  group = str_to_lower(group))

# makes everything lowercase in group variable
# modifies an already existing variable, overwrites and does not change anything else

mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight, 
  wt_gain_squared = wt_gain ^ 2,
  group = str_to_lower(group))

# modifies and creates variables at same time 
```

In the pups dataset, create a variable that subtracts 7 from PD walk 

```{r}
mutate(pups_df, pd_walk_minus_7 = pd_walk - 7)
```

## Arrange the data 

```{r}
arrange(litters_df, pups_born_alive)
## rearrange so that they are in order of pups born alive 3 -> 4 -> 5 etc, missing values always go to bottom 

arrange(litters_df, desc(pups_born_alive)) 
## arrange in reverse order, descending order 
```

## Let's do multiple steps 

Do NOT do this: too many variables in environment. do not want to create sub datasets 
```{r}
litters_df = read_csv("data_import_examples/FAS_litters.csv", na = c("", "NA", "."))
litters_rename_df = janitor::clean_names(litters_df)
litters_with_gd_df = select(litters_rename_df, group, starts_with("gd"))
litters_no_na_df = drop_na(litters_with_gd_df)
litters_wt_gain_df = mutate(litters_no_na_df, wt_gain = gd18_weight - gd0_weight)
```

It is even worse to nest. Very confusing. 

Instead: 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("", "NA", ".")) |> 
  janitor:: clean_names() |> 
  select(group, starts_with("gd")) |> 
  drop_na() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight, 
    group = str_to_lower(group)
  )
```

Load pups, clean names, drop missing, keep litter number and pd variables, add pd walk - 7 

```{r}
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", skip = 3, na = c("", "NA", ".")) |> 
  janitor::clean_names() |> 
  drop_na() |> 
  select(litter_number, starts_with("pd")) |> 
  mutate(
    pd_walk_minus_7 = pd_walk - 7
  )
```

Sometimes the first thing isn't a dataframe 
lm = linear model
Take this thing, have to use data = _ as a placeholder for litters_df, you could also do litters_df instead of _

```{r}
litters_df |> 
  filter(group %in% c("con7", "con8")) |> 
  lm(gd18_weight ~ gd0_weight, data = _)
```

Select vs. Pull 
```{r}
litters_df |> 
  select(group)

#Select will always give you a dataframe. 

litters_df |> 
  pull(group)

litters_df |> 
  filter(group %in% c("Con7", "Con8")) |> 
  pull(group)

# this creates a vector, not a df. 
# Almost never do this 
```

Never do $

```{r}
litters_df$group
# does same thing as pull 
# do not do this, keep stuff in dataframe, gets confusing. 
# Better to mutate within the dataframe 
```

Pipes and Viewing 

```{r}
litters_df |> 
  filter(group %in% c("Con7", "Con8")) |> 
 # view()

#This will pop out a separate window to see df
# DELETE this step, do not include in Rmd code. can do it along the way to double check. 
# Can also do in console 
```






