---
title: "Data Manipulation"
author: "Emma"
date: "2026-09-24"
output: github_document
---

This file is for doing data manipulation. 

```{r message = F}
library(tidyverse)
```

Load the litters and pups datasets. 

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", 
           na = c("", "NA", "."))

litters_df = janitor::clean_names(litters_df)

pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", 
           skip = 3,
           na = c("", "NA", "."))

pups_df = janitor::clean_names(pups_df)
```

## Select some variables 
Which data frame, and which variable columns are you selecting 
can list the columns, or do this through that (:), the - will do everything except what you -
GROUP = group renames that variable by itself (will also put it first), the everything says keep everything else

```{r}
select(litters_df, group, litter_number)

select(litters_df, group:gd_of_birth)

select(litters_df, -pups_survive)

select(litters_df, -group)

select(litters_df, starts_with("gd"))

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP = group, everything())

select(litters_df, -starts_with("gd"))
```

A couple alternatives 
```{r}
rename(litters_df, GROUP = group)
#Same thing as above except you don't have to tell it to keep everything else 

relocate(litters_df, litter_number, gd0_weight)
# relocate moves litter_number first, gd0_weight second (based off list order)
```


In the pups data, select litter number, sex, and PD ears: 

```{r}
select(pups_df, litter_number, sex, pd_ears)
```

## Filter some rows 
Which rows do I need to keep? Filter and keep all the litters where x, y (something inside your data set)
```{r}
filter(litters_df, gd_of_birth == 19)
# Asking a question with ==, which of gd_of_birth is equal to 20? 

filter(litters_df, pups_born_alive <= 5)
# less than or equal to 5 

filter(litters_df, pups_born_alive != 6)
# not equal to 6 (gets rid of 6)

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive < 7)
# going to keep both of those two options only, can filter for multiple things at same time

filter(litters_df, pups_born_alive < 4  | pups_born_alive > 8)
#this means or |

drop_na(litters_df)
# gets rid of all rows you have missing value 

drop_na(litters_df, gd0_weight)
# drop na for all in this specific column
```

In the pups dataset, keep only rows where PD walk is less than 11 and sex is 2 

```{r}
filter(pups_df, pd_walk < 11, sex == 2)
```

#Mutate to change some columns 

Let's add and change some columns 

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)
# creates a column in dataset, what is weight gain from beginning of pregnancy to right before birth
# puts variable at tail end
# does not change anything else 

mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight, 
  group = str_to_lower(group))

# makes everything lowercase in group variable
# modifies an already existing variable, overwrites and does not change anything else

mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight, 
  wt_gain_squared = wt_gain ^ 2,
  group = str_to_lower(group))

# modifies and creates variables at same time 
```

In the pups dataset, create a variable that subtracts 7 from PD walk 

```{r}
mutate(pups_df, pd_walk_minus_7 = pd_walk - 7)
```





