---
title: "Tidy Data Notes"
author: "Emma"
date: "2026-09-29"
output: github_document
---
##Notes 
```{r message = F}
library(tidyverse)
```


```{r}
data("mtcars")
```

```{r}
mtcars$mpg

mtcars$mpg = 1 
# this pulls something out and manipulates the whole data set. don't do this 
```

```{r}
pull(mtcars, mpg) 
# if you have take something out, use pull function like this because it won't let you edit whole dataset
```

```{r}
mtcars |> 
  select(mpg:disp) 

#be careful when loading multiple libraries. they could use the same words for different functions, and R will remember the most recently loaded library. to avoid this, add which package from the library you're using: 

mtcars |> 
  dplyr::select(mpg:disp)
```

#Tidy data 
Rules for tidy data: 
* data tables have an implied structure which the "tidy data" framework makes explicit 
** Columns are variables 
** Rows are observations 
** every value has a cell 

Why tidy your data?
* consistent data structures will simplify your thought process 
** especially true if you use tools designed to tidy data 
* data written for computers is easier to work with 

Not all data are tidy: 
* columns are values not variable names 
* single columns contain multiple variables 
* data stored in multiple tables 
WATCH OUT FOR THESE! 

Relational data 
* data spread across tables with defined relations 
* variables used to define these relations are *keys*
* tables are combined by *joins* 

Join types 
* joining datasets x and y 
** mostly use left joins and full joins (both outer joins)

Key functions 
* For tidying single tables 
** pivot_longer --> easier to read as computer 
** separate (separate column into two separate columns) 

* for untidying single tables:
** pivot_wider --> easier to read as human

* For combining multiple tables 
** bind_rows
** *_join --> whichever join you're doing 


